State of the Practice for Lattice Boltzmann Method Software

### 2022-01-10 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revised order of sections, added more explicit references to the RQs

\begin{enumerate}
	\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What LBM
	software projects exist, with the constraint that the source code must be
	available for all identified projects?
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Which
	of the projects identified in \rqref{RQ_WhatProjects} follow current best
	practices, based on evidence found by experimenting with the software and
	searching the artifacts available in each project's repository?
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
	similar is the list of top projects identified in \rqref{RQ_HighestQuality}
	to the most popular projects, as viewed by the scientific community?
    \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
	do LBM projects compare to research software in general with respect to
	the artifacts present in their repositories?
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:]
	How do LBM projects compare to research software in general with
	respect to the use of tools for:
	\begin{enumerate} 
		\item [\rqref{RQ_CompareToolsProjMngmnt}.a] development; and,
		\item [\rqref{RQ_CompareToolsProjMngmnt}.b] project management?
	\end{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
	the pain points for developers working on LBM software projects? 
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
	do the pain points of developers from LBM compare to the pain points
	for research software in general?
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For LBM
	developers:
	\begin{enumerate} 
		\item [\rqref{RQ_Concerns}.a] what software development concerns exist; and,
		\item [\rqref{RQ_Concerns}.b] what specific actions are taken to address
		these concerns?
	\end{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareConcerns}:]
	Compared to research software in general, how is LBM software (as answered
	in \rqref{RQ_Concerns}) different with respect to:
	\begin{enumerate} 
		\item [\rqref{RQ_CompareConcerns}.a] software development concerns?; and,
		\item [\rqref{RQ_CompareConcerns}.b] specific actions taken to address
		concerns?
	\end{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Future}:] What actions
	could be employed by future LBM developers to address the concerns
	identified in \rqref{RQ_Concerns} and the pain points found in
	\rqref{RQ_PainPoints}?

@@

We developed a methodology for evaluating the state of the practice of
scientific software \citep{SmithEtAl2021}.  This methodology can be instantiated
for a specific domain of scientific software, which in the current case is LBM
software.  The first section below describes the steps of the process we used to
select, measure and compare LBM software. This is followed by quality
definitions and how these qualities were assessed in our project. The rest of
the section provides the details of the methodology.
% an overview of the specifics of candidate software selection, the quantitative
% measurements, developer interviews, the ranking process via AHP and
% interaction with the Domain Expert.

\subsection{Process}

@@

The candidate software was found through search engine queries targeting
authoritative lists of software. We found LBM software listed on the websites
GitHub and swMATH, as well as through articles found in scholarly journals and
databases.  The Domain Expert (Section~\ref{sec_vet_software_list}) was also
engaged in selecting the candidate software.  \wss{Packages that fit our
criteria but were found after data collection and analysis was conducted, such
as Musubi \citep{HasertEtAl2014}, will need to be added to future SOP
assessments.}

@@

To reduce the number of members in the candidate software list to a manageable
size, the following filters were applied. The filters were applied in the
priority order listed.

@@

The measurement template is found in \citet{SmithEtAl2021}.  This template is
used to track measurements and quality scores for all of the software packages
in the domain. For each software package, we fill-in the rows of the template.
This process takes about 2 hours per package, with a cap of 4 hours.  The time
constraint is necessary so that the work load for measuring a given domain is
feasible for a team as small as one, since we aim for a cap of 160 person hours
for the measurement phase \citet{SmithEtAl2021}.  An excerpt of the spreadsheet
is shown in Figure~\ref{measurement_template_image}.  A column is added to this
template for each of the 24 software package measured.

@@

The Analytical Hierarchy Process (AHP) is a decision-making technique that is
used to compare multiple options by multiple criteria. In our work AHP was used
for comparing and ranking the LBM software packages using the overall impression
quality scores that were gathered in the measurement template.  AHP performs a
pairwise analysis using a matrix and generates an overall score as well as
individual quality scores for each software package. \citep{SmithEtAl2016} shows
how AHP is applied to ranking software based on quality measures.

@@

We designed a list of 20 questions to guide our interviews, which can be found
in \citet{SmithEtAl2021}. Some questions are about the background of the
software, the development teams, the interviewees, and how they organize the
projects. We also ask about the developer's understandings of the users. Some
questions focus on the current and past difficulties, and the solutions the team
has found, or will try. We also discuss the importance and current situations of
documentation. A few questions are about specific software qualities, such as
maintainability, understandability, usability, and reproducibility. The
interviews are semi-structured based on the question list; we ask follow-up
questions when necessary. Based on our experience, the interviewees usually
bring up some exciting ideas that we did not expect, and it is worth expanding
on these topics.

@@

The Domain Expert is an important member of the state of the practice assessment
team. Pitfalls exist if non-experts attempt to acquire an authoritative list of
software, or try to definitively rank the software. Non-experts have the problem
that they can only rely on information available on-line, which has the
following drawbacks:
\begin{inparaenum}[i)]
  \item the on-line resources could have false or inaccurate information; and,
  \item the on-line resources could leave out relevant information that is so
in-grained with experts that nobody thinks to explicitly record it.
\end{inparaenum}
For the current assessment, our Domain Expert (and paper co-author) is Dr.\
Zahra Motamed, Assistant Professor of Mechanical Engineering at McMaster
University, Hamilton, Ontario, Canada.

@@

Before discussing quality differences between the packages and the results of
the AHP ranking, we will contrast the different packages based on features, as
shown in Table~\ref{tbl_features}. \wss{Add some discussion of some example
features.  Add Non-Newtonian and Fluid Structure interaction.  Listed in
alphabetical order.}  The sections below summarize how the software packages
compare based on qualities, as measured by the measurement template.   The
results of the AHP ranking are also discussed.

@@

Table~\ref{repometrics} presents our LBM software package rankings along with
the repository ranking metrics of each software package. Nine packages do not
use GitHub, so they do not have a measure of repository stars. Looking at the
repository stars of the other 15 packages, we can observe a slight pattern where
packages that have been highly ranked by our assessment do have more stars than
lower ranked packages. The best ranked package (ESPResSo) has the second most
number of stars. The ninth ranked package (Sailfish) has the most number of
stars. Packages designated as lower quality often do not use GitHub or GitLab,
or have few stars. HemeLB only has 12 stars, but its location recently moved
from a different GitHub repository. Our assessment of this package might not be
accurate. The number of stars does not necessarily represent the perceptions of
the community, but for lack of an alternative measure we will use it this way.
Besides missing data, another threat to the validity of this comparison is the
varying ages of the repositories. Older packages have been able to accumulate
stars and watches for longer than newer packages. The true quality of new
packages may not be reflected in their stars and watches.

@@

\section{Qualitative Findings From Developer Interviews} \label{interviewresults}

This section presents qualitative findings on software qualities as gleaned from
interviewing 4 developers. The full interview questions are found in
\citet{SmithEtAl2021}.

@@

\section{Answers To Research Questions} \label{answersquestions}

This section answers the research questions listed in Section~\ref{secIntro}
using the quantitative data presented in Section~\ref{AHPresults}, qualitative
data presented in Section~\ref{interviewresults}, additional data from software
package repositories and artifacts, and domain expert feedback. Each subsection
answers one of the research questions. Software package artifacts are examined
in Section~\ref{artifacts}, tools are listed in Section~\ref{tools}, development
principles, processes, and methodologies are discussed in Section
~\ref{prinprocmeth}, development pain points are explored in Section
~\ref{painpoints}, and summary of current (Section~\ref{currentPractices}), best
(Section~\ref{bestPractices}) and future practices
(Section~\ref{futurePractices}). Finally, threats to the validity of this
assessment are noted in Section~\ref{threats}.

\subsection{What artifacts are present in current software projects?}
\label{artifacts}

The software packages were examined for the presence of artifacts, which were
then categorized by frequency. We have grouped them into common, less common,
and rare artifacts in Table~\ref{artifactspresent}. Common artifacts were found
in 16 to 24 ($>$63\%) of the software packages. Less common artifacts were found
in 8 to 15 (30-63\%) of the software packages. Rare artifacts were found in 1 to
7 ($<$30\%) of the software packages.

\subsection{What tools (development, dependencies, project management) are used
by current software packages?} \label{tools}

Software tools are used to support the development, verification, maintenance,
and evolution of software, software processes, and artifacts
\citep{ghezzi1991fundamentals}. Many tools are used by LBM software packages, as
summarized in Table~\ref{tbl_tools}.  The tools are subdivided into development
tools, dependencies, and project management tools.

\subsection{What principles, processes, and methodologies are used?}
\label{prinprocmeth}

Most of the software packages do not explicitly state in their artifacts the
motivations or design principles that were considered when developing the
software. One package, Sailfish, indicates in its artifacts that shortening the
development time was considered in early stages of design, with the developers
opting for using Python and CUDA/OpenCL to achieve this without sacrificing any
computational performance. The goals of that project are explicitly listed as
performance, scalability, agility and extendability, maintenance, and ease of
use. The project scored well in these categories during our assessment.

\subsection{What are the pain points?} \label{painpoints}

This subsection answers the research question: What are the pain points for
developers working on research software projects? What aspects of the existing
processes, methodologies and tools do they consider as potentially needing
improvement? How should processes, methodologies and tools be changed to improve
software development and software quality?  Developers were asked to comment on
obstacles in their development process, obstacles encountered by users, and
potential future obstacles.

@@

practices for the future.

>>>>

\begin{enumerate}
	\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What LBM
	software projects exist, with the constraint that the source code must be
	available for all identified projects?
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Which
	of the projects identified in \rqref{RQ_WhatProjects} follow current best
	practices, based on evidence found by experimenting with the software and
	searching the artifacts available in each project's repository?
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
	similar is the list of top projects identified in \rqref{RQ_HighestQuality}
	to the most popular projects, as viewed by the scientific community?
    \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
	do LBM projects compare to research software in general with respect to
	the artifacts present in their repositories?
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:]
	How do LBM projects compare to research software in general with
	respect to the use of tools for:
	\begin{enumerate} 
		\item [\rqref{RQ_CompareToolsProjMngmnt}.a] development; and,
		\item [\rqref{RQ_CompareToolsProjMngmnt}.b] project management?
	\end{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
	How do LBM projects compare to research software in general with respect to
	principles, processes and methodologies used?
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
	the pain points for developers working on LBM software projects? 
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
	do the pain points of developers from LBM compare to the pain points
	for research software in general?
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For LBM
	developers:
	\begin{enumerate} 
		\item [\rqref{RQ_Concerns}.a] what software development concerns exist; and,
		\item [\rqref{RQ_Concerns}.b] what specific practices are taken to
		address these concerns?
	\end{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
	What, if any, existing research software best practice, not currently used
	by LBM software, could address the concerns identified in
	\rqref{RQ_Concerns}).
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Future}:] What practices
	could be employed by future LBM developers to address the pain points found
	in \rqref{RQ_PainPoints} and the concerns identified in \rqref{RQ_Concerns}.a?

@@

We developed a methodology for evaluating the state of the practice of
scientific software \citep{SmithEtAl2021}.  The methodology aims to answer the
research questions posed in the previous section. This methodology can be
instantiated for a specific domain of scientific software, which in the current
case is LBM software.  The first section below describes the steps of the
process we used to select, measure and compare LBM software. This is followed by
quality definitions and how these qualities were assessed in our project. The
rest of the section provides the details of the methodology.
% an overview of the specifics of candidate software selection, the quantitative
% measurements, developer interviews, the ranking process via AHP and
% interaction with the Domain Expert.

\subsection{Overall Process}

@@

Part of how we answer \rqref{RQ_WhatProjects}.  The candidate software was found
through search engine queries targeting authoritative lists of software. We
found LBM software listed on the websites GitHub and swMATH, as well as through
articles found in scholarly journals and databases.  The Domain Expert
(Section~\ref{sec_vet_software_list}) was also engaged in selecting the
candidate software.  \wss{Packages that fit our criteria but were found after
data collection and analysis was conducted, such as Musubi
\citep{HasertEtAl2014}, will need to be added to future SOP assessments.}

@@

With previous section, how we answer \rqref{RQ_WhatProjects}.  Answer to
\rqref{RQ_WhatProjects} provided in this section.  To reduce the number of
members in the candidate software list to a manageable size, the following
filters were applied. The filters were applied in the priority order listed.

@@

Part of how we address \rqref{RQ_HighestQuality} and
\rqref{RQ_CompareArtifacts}.  Once have \rqref{RQ_HighestQuality} can combine
with estimate of popularity to answer \rqref{RQ_CompareHQ2Popular}.  The
measurement template is found in \citet{SmithEtAl2021}.  This template is used
to track measurements and quality scores for all of the software packages in the
domain. For each software package, we fill-in the rows of the template. This
process takes about 2 hours per package, with a cap of 4 hours.  The time
constraint is necessary so that the work load for measuring a given domain is
feasible for a team as small as one, since we aim for a cap of 160 person hours
for the measurement phase \citet{SmithEtAl2021}.  An excerpt of the spreadsheet
is shown in Figure~\ref{measurement_template_image}.  A column is added to this
template for each of the 24 software package measured.

@@

Part of how we address \rqref{RQ_HighestQuality}.  The Analytical Hierarchy
Process (AHP) is a decision-making technique that is used to compare multiple
options by multiple criteria. In our work AHP was used for comparing and ranking
the LBM software packages using the overall impression quality scores that were
gathered in the measurement template.  AHP performs a pairwise analysis using a
matrix and generates an overall score as well as individual quality scores for
each software package. \citep{SmithEtAl2016} shows how AHP is applied to ranking
software based on quality measures.

@@

Part of how we address \rqref{RQ_CompareToolsProjMngmnt},
\rqref{RQ_CompareMethodologies}, \rqref{RQ_PainPoints} (which enables
\rqref{RQ_ComparePainPoints}, and \rqref{RQ_Concerns} (which enables
\rqref{RQ_CompareConcerns}). We designed a list of 20 questions to guide our
interviews, which can be found in \citet{SmithEtAl2021}. Some questions are
about the background of the software, the development teams, the interviewees,
and how they organize the projects. We also ask about the developer's
understandings of the users. Some questions focus on the current and past
difficulties, and the solutions the team has found, or will try. We also discuss
the importance and current situations of documentation. A few questions are
about specific software qualities, such as maintainability, understandability,
usability, and reproducibility. The interviews are semi-structured based on the
question list; we ask follow-up questions when necessary. Based on our
experience, the interviewees usually bring up some exciting ideas that we did
not expect, and it is worth expanding on these topics.

@@

Part of how we address \rqref{RQ_WhatProjects} and \rqref{RQ_HighestQuality}.
\wss{Should ask Zahra about \rqref{RQ_CompareHQ2Popular}, see if he has anything
to add about \rqref{RQ_PainPoints}.} The Domain Expert is an important member of
the state of the practice assessment team. Pitfalls exist if non-experts attempt
to acquire an authoritative list of software, or try to definitively rank the
software. Non-experts have the problem that they can only rely on information
available on-line, which has the following drawbacks:
\begin{inparaenum}[i)]
  \item the on-line resources could have false or inaccurate information; and,
  \item the on-line resources could leave out relevant information that is so
in-grained with experts that nobody thinks to explicitly record it.
\end{inparaenum}
For the current assessment, our Domain Expert (and paper co-author) is Dr.\
Zahra Motamed, Assistant Professor of Mechanical Engineering at McMaster
University, Hamilton, Ontario, Canada.

@@

Answers \rqref{RQ_HighestQuality}.  Before discussing quality differences between the packages and the results of
the AHP ranking, we will contrast the different packages based on features, as
shown in Table~\ref{tbl_features}. \wss{Add some discussion of some example
features.  Add Non-Newtonian and Fluid Structure interaction.  Listed in
alphabetical order.}  The sections below summarize how the software packages
compare based on qualities, as measured by the measurement template.   The
results of the AHP ranking are also discussed.

@@

Answers \rqref{RQ_CompareHQ2Popular}. Table~\ref{repometrics} presents our LBM
software package rankings along with the repository ranking metrics of each
software package. Nine packages do not use GitHub, so they do not have a measure
of repository stars. Looking at the repository stars of the other 15 packages,
we can observe a slight pattern where packages that have been highly ranked by
our assessment do have more stars than lower ranked packages. The best ranked
package (ESPResSo) has the second most number of stars. The ninth ranked package
(Sailfish) has the most number of stars. Packages designated as lower quality
often do not use GitHub or GitLab, or have few stars. HemeLB only has 12 stars,
but its location recently moved from a different GitHub repository. Our
assessment of this package might not be accurate. The number of stars does not
necessarily represent the perceptions of the community, but for lack of an
alternative measure we will use it this way. Besides missing data, another
threat to the validity of this comparison is the varying ages of the
repositories. Older packages have been able to accumulate stars and watches for
longer than newer packages. The true quality of new packages may not be
reflected in their stars and watches.

@@

\section{Comparison of Artifacts to Research Software in General}
\label{Sec_CompareArtifacts}

Answers \rqref{RQ_CompareArtifacts}.  The software packages were examined for
the presence of artifacts, which were then categorized by frequency. We have
grouped them into common, less common, and rare artifacts in
Table~\ref{artifactspresent}. Common artifacts were found in 16 to 24 ($>$63\%)
of the software packages. Less common artifacts were found in 8 to 15 (30-63\%)
of the software packages. Rare artifacts were found in 1 to 7 ($<$30\%) of the
software packages.

\section{Comparison of Tool Usage to Research Software in General}
\label{Sec_CompareTools}

Answers \rqref{RQ_CompareToolsProjMngmnt}.  Software tools are used to support
the development, verification, maintenance, and evolution of software, software
processes, and artifacts \citep{ghezzi1991fundamentals}. Many tools are used by
LBM software packages, as summarized in Table~\ref{tbl_tools}.  The tools are
subdivided into development tools, dependencies, and project management tools.

\section{Comparison of Principles, Process and Methodologies to Research Software in General} \label{Sec_CompareMethodologies}

Answers \rqref{RQ_CompareMethodologies}.  Most of the software packages do not
explicitly state in their artifacts the motivations or design principles that
were considered when developing the software. One package, Sailfish, indicates
in its artifacts that shortening the development time was considered in early
stages of design, with the developers opting for using Python and CUDA/OpenCL to
achieve this without sacrificing any computational performance. The goals of
that project are explicitly listed as performance, scalability, agility and
extendability, maintenance, and ease of use. The project scored well in these
categories during our assessment.

\section{Developer Pain Points} \label{painpoints}

Answers \rqref{RQ_PainPoints} and \rqref{RQ_ComparePainPoints}.  This subsection
answers the research question: What are the pain points for developers working
on research software projects? What aspects of the existing processes,
methodologies and tools do they consider as potentially needing improvement? How
should processes, methodologies and tools be changed to improve software
development and software quality? Developers were asked to comment on obstacles
in their development process, obstacles encountered by users, and potential
future obstacles.

\section{LBM Developer Identified Concerns} \label{Sec_Concerns}

Answers \rqref{RQ_Concerns}.  This section presents qualitative findings on
software qualities as gleaned from interviewing 4 developers. The full interview
questions are found in \citet{SmithEtAl2021}.

@@

\section{Recommendations for Addressing Developer Concerns} \label{Sec_AddressConcerns}

Answers \rqref{RQ_Recommend}.  This section answers the research questions
listed in Section~\ref{secIntro} using the quantitative data presented in
Section~\ref{AHPresults}, qualitative data presented in
Section~\ref{interviewresults}, additional data from software package
repositories and artifacts, and domain expert feedback. Each subsection answers
one of the research questions. Software package artifacts are examined in
Section~\ref{artifacts}, tools are listed in Section~\ref{tools}, development
principles, processes, and methodologies are discussed in Section
~\ref{prinprocmeth}, development pain points are explored in Section
~\ref{painpoints}, and summary of current (Section~\ref{currentPractices}), best
(Section~\ref{bestPractices}) and future practices
(Section~\ref{futurePractices}). Finally, threats to the validity of this
assessment are noted in Section~\ref{threats}.

@@

Answers \rqref{RQ_Future}.  practices for the future.
